---
title: "#Crash course into statistics of metagenomic data: Multivariate analysis and generalized linear models of overdispersed count data"
output:
  github_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#Setting up the R-environment and compiling data

###Setting up needed packages

First we will install and load packages which are needed for analysis. Others are essential, viridis is used for getting colorblind friendly colors.
```{r}
#Install required libraries
#source('http://bioconductor.org/biocLite.R')
#biocLite('phyloseq')
#install.packages("vegan")
#install.packages("viridis")
#install.packages("MASS")

#Load libraries
library(phyloseq)
library(viridis)
library(vegan)
library(MASS)
library(ggplot2)
```

###Normalizing data

Count data of metagenomes is almost always normalized to either library sizes or 16S rRNA gene counts to get rid off variation that is caused due to differences in library sizes or the numbers of bacteria. In our case, we have a lot of non-bacterial DNA in our samples, and we are quatifying genes which are found exclusively in bacteria. The amount of bacterial DNA also varies between samples. This is why we will normalize our data to 16S counts.


```{r}
#Load in the ARG mapping result
ARG_bt<-as.matrix(read.table("ARG_genemat.txt", fill= 1, header= T, row.names = 1, check.names = F))

#Let's remove a gene which is not relevant for us

ARG_bt<-ARG_bt[grep('oqxB', row.names(ARG_bt), invert = TRUE),]

#Read in metadata
sample_data<-read.table("metadata.txt", sep="\t")

#Normalize to 16S counts
ARG_bt_norm<-t(t(ARG_bt)/sample_data$SSU_SUM)

#Check the division by using a random column and row
identical(ARG_bt[1083, 10]/sample_data$SSU_SUM[10], ARG_bt_norm[1083, 10])

```

###Phyloseq is handy for organizing and accessing data with counts, taxonomy and metadata files

Let's use Phyloseq to organize our data. Phyloseq is good for making sure your columns and rows are always in the right order and it makes accessing and splitting data faster.

```{r}
#Make Phyloseq object with sample data and otu table. 
baby_ARG_PHY<-phyloseq(otu_table(ARG_bt_norm, taxa_are_rows = TRUE), sample_data(sample_data))

#Look at the sample sums
sample_sums(baby_ARG_PHY)

#Check the sample data variable DIET
sample_data(baby_ARG_PHY)$DIET
```

##Multivariate analysis

Often the research question is if there are differences in all genes between samples that can be explained by an explanatory variable or variables. This is when multivariate analysis methods come in handy.

Multivariate methods can be used to calculate distances between all genes/taxa in samples using dissimilarity/similarity indexes and ordinating the samples in a two-dimensional space for visualization. One of the most commonly used distances in ecology is the Bray-Curtis dissimilarity index. There are several distances such as Euclidean, Jaccard and Horn-Morisita. However, it is best to use a method which does not take zero values into account for metagenomic data.

Ordination methods are used to ordinate samples in relation to each other in multiple dimensions. There are several ordination methods such as NDMS, PCA and PCoA. We will use PCoA. Tranforming your data is often needed before this, but we will skip it for the figure. The first two dimensions or axes of the ordination can be used to plot the location of samples in relation to each other in a two-dimensional space.

Phyloseq has vegan built-in so we can ordinate distances using Phyloseq and plot the ordination.

Phyloseq has vegan built-in so we can ordinate distances using Phyloseq and plot the ordination.

###Ordination of samples using PCoAs and Bray-Curtis dissimilarity index

```{r}
#Ordinate using Phyloseq's ordinate command
baby_ARG_PHY_ord<-ordinate(baby_ARG_PHY, "PCoA", "bray")

#Plot ordination and color by Formula
p<-plot_ordination(baby_ARG_PHY, baby_ARG_PHY_ord, color="DIET")
p+scale_color_viridis(discrete = T) + geom_point(size=3) + theme_minimal()

```

We can see that there are more purple points on the left. Let's test if the difference we see visually is also statistically significant using the 'adonis' function from the vegan package. Now we will also square root transform the count table which can be accessed by using the function 'otu_table' from the phyloseq package.

```{r}
#We use a square root transformation for the data to even out variation
ARG_dist<-vegdist(sqrt(t((otu_table(baby_ARG_PHY)))), dist = "bray")

adonis(ARG_dist~DIET, data=data.frame(sample_data(baby_ARG_PHY), permutations = 9999))
```

We can conclude that is some diffference in the grouping of genes based on the explanatory variables. (The separation is borderline significant using alpha of 0.05). Adonis calculates the result using permutations, so there can be some variation in the p-values and model estimates with lower numbers of permutations. If we would like to know which individual genes are different we could use the R-package DeSEQ to investigate fold changes between the two groups or fit genes as vectors in the ordination using 'envfit' from vegan package.


##Analysis of diversity

The diversity of bacteria or genes can differ between groups and is often interesting to study. Again, there are several diversity indeces, but in most cases it is best to use an indes which takes both evenness and the number of different taxa or genes into account when analysing metagenomes.

Phyloseq is able to plot diversities so we will use the 'plot_richness' function from phyloseq.

```{r}
p<-plot_richness(baby_ARG_PHY, measures = c("Shannon", "Simpson"), x = "DIET", color="DIET")
p+geom_boxplot() +
scale_color_viridis(discrete=TRUE) +
geom_point(size=3)
```

Again, it looks like there might be a difference between the groups, but lets check if it's significant or not using ANOVA, which compares the variances between the two groups.

```{r}
shannon<-aov(diversity(t(otu_table(baby_ARG_PHY)), index = "shannon")~sample_data(baby_ARG_PHY)$DIET, 
             data=as.data.frame(otu_table(baby_ARG_PHY)))
TukeyHSD(shannon)

simpson<-aov(diversity(t(otu_table(baby_ARG_PHY)), index = "simpson")~sample_data(baby_ARG_PHY)$DIET, 
             data=as.data.frame(otu_table(baby_ARG_PHY)))
TukeyHSD(simpson)

```

It turns out that the difference is not significant using Simpson diversity but is with Shannon.

##Statistical analysis of overdispersed count data 

Many read based metagenomic analyses produce outputs in table formats with counts. During the course we have analysed our data and counted ARGs in each sample by mapping them to an ARG database. We need to decide which statistical methods are most appropriate for out data, in order to analyse the effects of the 'DIET' explanatory variable on the ARG counts.

In traditional statistics, count data is usually modelled using Poisson or binomial distibutions. Poisson is a discrete probability distribution that expresses the probability of a given number of events occurring in a fixed interval of time or space IF these events occur with a known constant rate and independently of the time since the last event. For instance, a person keeping track of the amount of mail they receive each day may notice that they receive an average number of 4 letters per day. If receiving any particular piece of mail does not affect the arrival times of future pieces of mail, i.e., if pieces of mail from a wide range of sources arrive independently of one another, then a reasonable assumption is that the number of pieces of mail received in a day obeys a Poisson distribution. However, more often than not this is not the case for real life events and biological data.

The starting point for count data is a GLM (generalized linear model) with Poisson-distributed errors, but not all count data meet the assumptions of the Poisson distribution. Poisson distribution requires that the variance is NOT larger than the mean (overdispersed) and that there aren't more zeros than expected in the data (zero inflation).

The binomial distribution is frequently used to model the number of successes in a sample of size n drawn with replacement from a population of size N. If the sampling is carried out without replacement, the draws are not independent and so the resulting distribution is a hypergeometric distribution, not a binomial one. However, for N much larger than n, the binomial distribution remains a good approximation, and is widely used. As with the Poisson distribution, binomial distribution's assumptions include that there is no overdispersion and zero inflation.

Some also often use normally distributed models and log transfor their data, but in most cases the models will end up beign less than optimal. Why not use the right model for your data instead of compromizing?

###Checking which model is best - Poisson, Quasipoisson or Negative Binomial
```{r}
#Save the total sum of ARGs/average 16S rRNA gene copies to a dataframe for each sample and info of the DIET 
df<-data.frame(SUM=round(mean(sample_data(baby_ARG_PHY)$SSU_SUM)*(sample_sums(baby_ARG_PHY))),
               DIET=as.factor(sample_data(baby_ARG_PHY)$DIET))

#First we can plot the data using boxplots, which plot the interquartile ranges and median
plot(df$SUM~df$DIET)

```

You can see that there might be a difference between 'n' and 'y', but there is large variation between the samples in the 'y' group which has the bigger median. This already gives us an idication that there is a lot of variation in the data and that is definately not suitable for the Poisson model.

###Let's fit a linear model to the data

```{r}
model.ln<-lm(SUM~DIET, data=df)
summary(model.ln)
```
P-value looks good, but does the data fit the normality assumption

```{r}
shapiro.test(df$SUM)
plot(model.ln)
```

The Shapiro-Wilks test gives us a siginificant p-value, which means that the data significantly differs from the normal distribution. It also looks like we might have an s-shaped curve in the normality Q-Q plot.

###Let's try a linear model with a log transformation


```{r}
model.logln<-lm(log(SUM)~DIET, data=df)
summary(model.logln)
```

Again everything looks good but how's the normality.

```{r}
shapiro.test(log(df$SUM))
plot(model.logln)
```

Now the data fits the normality assumption.


###Let's try and fit a Poisson model to the data

```{r}
model.pois = glm(SUM ~ DIET , data = df, family = poisson)
summary(model.pois)

```

This looks pretty good if you just look at the the p-value, but the residual devience and AIC values, which indicate how well the model explains the variance in the data are big. This means that the model is not good at explaining the deviance.

###Let's check if the deviance is above the 0.95 critical value for the chi squared thest with 9 degrees of freedom

```{r}
qchisq(0.95, df.residual(model.pois))#Critical value

#Actual deviance for the Poisson model
deviance(model.pois)

#The deviance is much larger than 15.5

#Let's calculate the residual deviance also using the Pearson chi-squared test
pr <- residuals(model.pois,"pearson")

sum(pr^2)

#The deviance is again a lot higher that the critical value for the chi squared test


#Now if we assume that the variance is proportional rather than equal to the mean,
#we can use a Poisson model with the quasipoisson extension. The scale parameter φ can be estimated by dividing Pearson's chi-squared by its d.f.:
phi <- sum(pr^2)/df.residual(model.pois)

round(c(phi,sqrt(phi)),4)

#We see that the variance is 4914.6439 times larger than the mean. 
#This means that we should adjust the standard errors multiplying by 70.1045, the square root of 4914.6439

#R can do this calculation for us if we use the quasipoisson family:


model.qpois<-glm(SUM~DIET, data=df, family="quasipoisson")

summary(model.qpois)

#The estimates are exactly the same as before, but the standard
#errors are about 70X larger and the p-value is larger. 
#We can verify this fact easily. First we write a useful function to extract standard errors and then use it on our fits:

se <- function(model) sqrt(diag(vcov(model)))

round(data.frame(poisson=coef(model.pois), 
                 qpoisson=coef(model.qpois), 
                 se.poisson=se(model.pois), 
                 se.qpoisson=se(model.qpois), 
                 ratio=se(model.qpois)/se(model.pois)), 4)

```

#Negative binomial distribution

Negative binomial distributions are an other option for overdispersed data. Negative binomial models are a bit more advanced in modelling the deviance.

```{r}
model.nb <- glm.nb(SUM~DIET, data=df)

summary(model.nb)


#R's theta is the precision of the multiplicative random effect, and corresponds to 1/σ2 (σ2 = variance) in the notes. 
#The estimate corresponds to an estimated variance of 0.798 and is highly significant.

1/model.nb$theta

#0.7980112

#Let's estimate the dispersion in the negative binomial model
-2*(logLik(model.pois)-logLik(model.nb))


#Let's compare the results from linear models, Poisson, quasipoisson and negative binomial models

data.frame(linear=coef(model.ln),
                 loglinear=exp(coef(model.logln)),
                 poisson=exp(coef(model.pois)),
                 qpoisson=exp(coef(model.qpois)),
                 negbin=exp(coef(model.nb)),
                 se.ln=(se(model.ln)),
                 se.logln=exp(se(model.logln)),
                 se.pois=exp(se(model.pois)),
                 se.qpoisson=exp(se(model.qpois)),
                 se.negbin=exp(se(model.nb)))

#Now if we remember the 95% cutoff from chi-squared test was 15.50731
deviance(model.ln)
#We have 729799086 deviance in the linear model
deviance(model.logln)
#We have 15.52274 deviance in the log linear model
deviance(model.pois)
#We have 51331.7 deviance in the Poisson model
deviance(model.qpois)
#The deviance is the same in the quasipoisson which estimates the deviance in the confidence intervals
deviance(model.nb)
#The deviance in the negative binomial model is 11.25934,
#which is below the cutoff value which indicates that the negative binomial model is a good fit for the data.


```


###Checking variances between models



The over-dispersed Poisson (quasipoisson) and negative binomial models have different variance functions. One way to check which one may be more appropriate is to create groups based on the linear predictor, compute the mean and variance for each group, and finally plot the mean-variance relationship. Unfortunately, our data doesn't have enough data points to be able to get a nice look at the variance functions, but let's look at the variance plot produced by Univ. of Princeton people.

https://data.princeton.edu/wws509/r/overdispersion


The graph plots the mean versus the variance and overlays the curves corresponding to the over-dispersed Poisson model, where the variance is φμ, and the negative binomial model, where the variance is μ(1+μσ2).

The Poisson variance function does a pretty good job for the bulk of most data, but fails to capture the high variances of the most productive scholars. The negative binomial variance function is not too different but, being a quadratic, can rise faster and does a better job at the high end. We conclude that the negative binomial model provides a better description of the data than the over-dispersed Poisson model.




