---
title: "Humann2"
author: "Tommi"
output:
  github_document: default
---

#Humann2 analysis and visualization in R

##Set your working directory to where you have your data on your own computer (and install) and load needed libraries

```{r}
setwd("/Users/kparnane/Documents/MetagenomeCourse2019/R_for_Humann2/")
#install.packages("tidyverse")
#install.packages("vegan")
#install.packages("broom")

library(tidyverse)
library(vegan)
library(broom)
```


##Read in the normalized pathways file

```{r}

humann2_pathways <- as.data.frame(read_tsv("pathways_norm.tsv"))
```


##Re-format the data frame from wide to long format

tidyverse is a really handy package for data wrangling and reformatting tables. This block is commented for what each line does. You can check the help page for each command by typing ? and the command's name for exaple ?rename for the next blocks of code

```{r}
humann2_pathways_long <- 
  # %>% is used to pipe the output to the next command
  humann2_pathways %>% 
  #Rename  `# Pathway` as pathway
  rename(pathway = `# Pathway`) %>%
  #Gather cmp by pathway and sampleID
  gather(sampleID, cpm, -pathway) %>% 
  #Separate by sampleID and drop any extra values without warning
  separate(sampleID, "sampleID", sep = "_", extra = "drop") %>% 
  #Separate pathways from organisms using |
  separate(pathway, c("pathway", "organism"), sep = "\\|", fill = "right")

```

#Continue processing the humann2 output
```{r}
# generate pathway table with no organism stratifications
humann2_pathways_no_stratifications_long <-
  humann2_pathways_long %>%
  filter(is.na(organism)) %>%
  select(-organism) %>%
  filter(!(grepl("^UN", pathway))) 

# Compute pathway alpha divertities per sample
humann2_pathways_no_stratifications_long %>%
  group_by(sampleID) %>%
  summarise(shannons_div = vegan::diversity(cpm),
            num_pathways = sum(cpm>0)) 
# .. continue with any statistical comparisons etc.

# Work with organism level stratifications
humann2_pathway_stratifications_long <- 
  humann2_pathways_long %>%
  filter(!(is.na(organism))) %>%
  filter(!(grepl("^UN", pathway))) 

# number of organisms per pathway
humann2_organisms_per_pathway <- 
  humann2_pathway_stratifications_long %>%
  group_by(pathway) %>%
  summarise(num_organisms = length(unique(organism)))

# average contributional alpha diverity (Gini simpson diversity) per pathway
humann2_pathway_alpha_div <- 
  humann2_pathway_stratifications_long %>%
  filter(cpm > 0) %>%
  group_by(pathway, sampleID) %>%
  summarise(alpha_div = vegan::diversity(cpm, index = "simpson")) %>%
  group_by(pathway) %>%
  summarise(mean_alpha_div = mean(alpha_div),
            median_alpha_dv = median(alpha_div)) %>%
  arrange(-mean_alpha_div)

head(humann2_pathway_alpha_div)

# add number of samples where pathway present
humann2_pathway_alpha_div_with_n_samples <- 
  humann2_pathways_no_stratifications_long %>%
  filter(cpm > 0) %>%
  group_by(pathway) %>%
  summarise(n_samples = n()) %>%
  left_join(humann2_pathway_alpha_div) %>%
  left_join(humann2_organisms_per_pathway)

humann2_pathway_alpha_div_with_n_samples %>%
  filter(n_samples == 10) %>%
  arrange(mean_alpha_div) %>%
  print(n = 20)
# PWY0-162 as an example of low diversity pathway

humann2_pathway_alpha_div_with_n_samples %>%
  filter(n_samples == 10) %>%
  arrange(-mean_alpha_div)
# PWY-7219 as an example of high divertsity pathway
```

